---
title: "gusto_stress_data_wrangle"
author: "Fran Querdasi"
date: "4/30/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup
## load libraries
```{r}
library(tidyverse)
library(compareDF)
```

# Data Wrangling
## Read in microbiome metadata files
```{r}
#read in metadata files
long_m_file <- "../../data/microbiome/GUSTO_106subjects_16Sdata_longitudinal/original_files/Sample_IDs.txt"
cs_m_file <- "../../data/microbiome/M24_M48/original_files/SampleID_mapping.txt"

long_meta <-
  read_tsv(long_m_file)

cs_metadata <-
  read_tsv(cs_m_file)
```
## Join the cross-sectional and longitudinal microbiome M24 datasets' metadata together
```{r}
#rename gusto ID variable in cross-sectional metadata to be same as longitudinal metadata
cs_metadata <-
  cs_metadata %>% 
  dplyr::rename(GUSTO_id = GUSTOid, '#SampleID' = '#Sample_ID')

#create datasets with just 24 month samples for both cross-sectional and longitudinal 
cs_metadata_24 <-
  cs_metadata %>% 
  filter(Timepoint_months == 24) %>% 
  dplyr::select(-Sequencing_batch)

long_metadata_24 <-
  long_meta %>% 
  filter(Timepoint_months == 24)

#join cross-sectional and longitudinal 24M samples
metadata_24 <-
  cs_metadata_24 %>% 
  full_join(long_metadata_24, by = "GUSTO_id")
```

## Create datasets with each adversity exposure data and M24, M48 metadata
```{r}
#read in adversity data
postnatal_stress_file <- "../../data/stress/GUSTO_child_stress_Y4_20210528.csv"

ctq_file <- "../../data/mother_childhood/GUSTO_M54_CTQ_scores_revised_20230109.csv"

stai_file <- "../../data/mh_trajectories/GUSTO_BDISTAI_imputed_scores_20210507.csv"

postnatal_dat <-
  read_csv(postnatal_stress_file)

ctq_dat <-
  read_csv(ctq_file)

stai_dat <-
  read_csv(stai_file)

#select variables of interest for each adversity exposure 
#prenatal 
stai_ids <-
  stai_dat %>% 
  filter(!is.na(STAI_prorated_state_pw26)) %>% 
  dplyr::select(subjid, STAI_prorated_state_pw26)

#preconception
ctq_ids <-
  ctq_dat %>% 
  filter(!is.na(ctq_total_score_25it)) #only take samples with data

#postnatal
postnatal_ids <-
  postnatal_dat %>% 
  dplyr::select(
    subjid,
    les_child_neg_events_total_Y04,
    leq_neg_events_total_Y04,
    leq_b2_events_total_Y04,
    leq_b2_neg_events_total_Y04,
    child_neg_events_sum
  )

#select samples from microbiome metadata that were collected at M24
mb_ids_M24 <-
  metadata_24 %>% 
  mutate(timepoint = 24) %>% 
  dplyr::select(GUSTO_id, timepoint) %>% 
  dplyr::rename(subjid=GUSTO_id)

#select samples form microbiome metadata that were collected at M48
mb_ids_M48 <-
  cs_metadata %>% 
  mutate(timepoint=48) %>% 
  filter(Timepoint_months == 48) %>% 
  dplyr::select(GUSTO_id, timepoint) %>% 
  dplyr::rename(subjid=GUSTO_id)
```

## Merge 24M microbimoe metadata and adversity data
```{r}
M24_all <-
  postnatal_ids %>% 
  full_join(mb_ids_M24, by = "subjid") %>% 
  full_join(stai_ids, by = "subjid") %>% 
  full_join(ctq_ids, by = "subjid")
```

## Merge M48 microbimoe metadata and adversity data
```{r}
M48_all <-
  postnatal_ids %>% 
  full_join(mb_ids_M48, by = "subjid") %>% 
  full_join(stai_ids, by = "subjid") %>% 
  full_join(ctq_ids, by = "subjid")
```

## Add ethnicity variable
```{r}
#read in BDI file -- contains mom ethnicity
bdi_orig_f <- "../../data/pregnancy_mh/STAI_BDI_FormA515_20201218.xlsx"

mom_ethn <-
  readxl::read_xlsx(bdi_orig_f)

#select mom ethnicity values
mom_ethn <-
  mom_ethn %>% 
  dplyr::select(
    subjid=SubjectID,
    mother_ethnicity
  ) 

M24_all <-
  M24_all %>% 
  left_join(mom_ethn, by = "subjid")

M48_all <-
  M48_all %>% 
  left_join(mom_ethn, by = "subjid")
```


## Generate datasets combining continuous adversity data, microbiome metadata, and ethnicity 
```{r}
#filter for IDs that have microbiome data
M24_mb <-
  M24_all %>% 
  filter(!is.na(timepoint))

M48_mb <-
  M48_all %>% 
  filter(!is.na(timepoint))

#create #SampleID variable for merging with alpha diversity data
M24_mb <-
  M24_mb %>%
  mutate(month = "M") %>% 
  unite('#SampleID_p', subjid, timepoint, sep= "-", remove=FALSE) %>% 
  unite('#SampleID', '#SampleID_p', month, sep= "", remove=TRUE)

M48_mb <-
  M48_mb %>%
  mutate(month = "M") %>% 
  unite('#SampleID_p', subjid, timepoint, sep= "-", remove=FALSE) %>% 
  unite('#SampleID', '#SampleID_p', month, sep= "", remove=TRUE)
```

# Write datasets for use in main analysis script
```{r}
M24_output_path <- "../../data/analysis_datasets/fran/gusto_M24cs_stress_data_ctqrev.csv"
M48_output_path <- "../../data/analysis_datasets/fran/gusto_M48_stress_data.csv"

write_csv(M24_mb, M24_output_path)
write_csv(M48_mb, M48_output_path)
```

