---
title: "gusto_metadata_wrangle"
author: "Fran Querdasi"
date: "5/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup
## Load libraries
```{r, message=FALSE}
library(tidyverse)
```

## Import data
**Note:** Filepaths (this chunk) are not shown
```{r, echo=FALSE, message=FALSE}
#filepaths 
postnatal_stress_file <- "../../data/stress/GUSTO_child_stress_Y4_20210528.csv"
prenatal_stress_file <- "../../data/mh_trajectories/GUSTO_BDISTAI_imputed_scores_20210507.csv"
mom_childhood_stress_file <- "../../data/mother_childhood/GUSTO_M54_CTQ_scores_revised_20230109.csv"
ethnicity_file <- "../../data/pregnancy_mh/STAI_BDI_FormA515_20201218.xlsx"
infant_sex_file <- "../../data/general_covariates/infant_sex.xlsx"
metadata_file <- "../../data/microbiome/M24_M48/original_files/SampleID_mapping.txt"
income_file <- "../../data/general_covariates/GUSTO_income_clean_20210702.csv"
delivery_mode_file <- "../../data/mb_covariates/GUSTO_deliverymode_cleaned_20210623.csv"
breastfeeding_file <- "../../data/mb_covariates/GUSTO_breastfeeding_clean_20210623.csv"
biotics_file <- "../../data/mb_covariates/GUSTO_M24_biotics_clean_20220218.csv"
gestage_file <- "../../data/general_covariates/GA_weight_Ages_at_measures_FormA515_20210719.xlsx"
cbcl_age2_file <- "../../data/child_beh/GUSTO_CBCL_scores_M24_20220516.csv"
cbcl_age4_file <- "../../data/child_beh/GUSTO_CBCL_scores_Y04_20220516.csv"
age_file <- "../../data/general_covariates/GUSTO_age_stoolproxy_20220615.csv"
diet_file <- "../../data/mb_covariates/diet_covariates.csv"
dqi_file <- "../../data/mb_covariates/dqi_variety_scores.csv"


#read in dataframes
post_stress <-
  read_csv(postnatal_stress_file)  ##fran this doesn't work for me

pre_stress <-
  read_csv(prenatal_stress_file) %>% 
  dplyr::select(
    subjid,
    STAI_prorated_state_pw26,
    stai_state_above_clin_cutoff_pw26
  )

m_ch_stress <-
  read_csv(mom_childhood_stress_file) %>% 
  dplyr::select(
    subjid,
    maltreated_mod_rev,
    contains("score")
  )

ethnicity <-
  readxl::read_xlsx(ethnicity_file) %>% 
  dplyr::select(
    subjid = SubjectID,
    mother_ethnicity
  )

sex <-
  readxl::read_xlsx(infant_sex_file) %>% 
  dplyr::rename(subjid=SubjectID) #0 = female, 1 = male

metadata <-
  read_tsv(metadata_file) %>% 
  dplyr::rename(
    '#SampleID' = '#Sample_ID',
    subjid=GUSTOid) #rename ID variable for joining with other data

income <-read_csv(income_file)

deliv_mode <-read_csv(delivery_mode_file)

bf <-read_csv(breastfeeding_file)

biotics <- read_csv(biotics_file)

gest_age <- readxl::read_xlsx(gestage_file) %>% 
  dplyr::select(
    subjid = SubjectID,
    GA
  )

beh2 <- read_csv(cbcl_age2_file) %>% 
  dplyr::select( #select only cbcl age 2 years variables that we will be examining 
    subjid,
    age_m24_cbcl = age_m24,
    cbcl_mom_fill_m24,
    cbclintprobtot_m24,
    cbcltotprobtot_m24,
    cbcldevprobtot_m24
  )

beh4 <- read_csv(cbcl_age4_file) %>% 
  dplyr::select( #select only cbcl age 4 years variables that we will be examining 
    subjid,
    age_y4_cbcl = age_y4,
    cbcl_mom_fill_y4,
    cbclintprobtot_y4,
    cbcltotprobtot_y4,
    cbcladhdprobtot_y4,
    cbclattprobtot_y4,
    cbclextprobtot_y4,
    cbclsleepprobtot_y4,
    cbcldevprobtot_y4
  )

diet <- read_csv(diet_file)

dqi_variety <- read_csv(dqi_file) %>% 
  dplyr::select(
    subjid,
    total_variety,
    foodgroup_variety,
    protein_variety
  )

dqi_variety$subjid <- str_sub(dqi_variety$subjid, end=9)
```
# Merge metadata
```{r, message=FALSE}
metadata_complete <- #only include metadata for IDs that have gut microbiome data, so left_join everything to gm metadata file
  metadata %>% 
  left_join(
    ethnicity, by = "subjid"
  ) %>% 
  left_join(
    sex, by = "subjid"
  ) %>% 
  left_join(
    m_ch_stress, by = "subjid"
  ) %>% 
  left_join(
    pre_stress, by = "subjid"
  ) %>% 
  left_join(
    post_stress, by = "subjid"
  ) %>% 
  left_join(
    income, by = "subjid"
  ) %>% 
  left_join(
    deliv_mode, by = "subjid"
  ) %>% 
  left_join(
    bf, by = "subjid"
  ) %>% 
  left_join(
    biotics, by = "subjid"
  ) %>% 
  left_join(
    gest_age, by = "subjid"
  ) %>% 
  left_join(
    beh2, by = "subjid"
  ) %>% 
  left_join(
    beh4, by = "subjid"
  ) %>% 
  left_join(
    read_csv(age_file), by = "subjid"
  ) %>% 
  left_join(
    diet, by = "subjid"
  ) %>% 
  left_join(
    dqi_variety, by = "subjid"
  )

```

# Clean up & add new variables
```{r}
metadata_final <- 
  metadata_complete %>% 
  dplyr::select(
    -person_filling_Y04, #not necssary to include, because it is "mother" for all who answered
    -leq_neg_events_total_Y04, #remove LEQ variables that are not being used
    -les_child_neg_events_total_Y04,
    -c_age_years_stoolproxy_y4,
    -contains("days")
  ) %>% 
  mutate(
    postnatal_stress = case_when( #NOTE that postnatal_stress var definition is DIFFERENT for 24M and 48M 
      Timepoint_months == 48 & child_neg_events_sum > 0 ~ 1, #postnatal_stress is 1 for M48 if any negative events endorsed on LEQ or any child events endorsed on LES
      Timepoint_months == 48 & child_neg_events_sum == 0 ~ 0,
      Timepoint_months == 24 & leq_b2_events_total_Y04 > 0 ~ 1, #postnatal_stress is 1 for M24 if any events endorsed before age 2 years
      Timepoint_months == 24 & leq_b2_events_total_Y04 == 0 ~ 0
    )
  ) %>% 
  dplyr::select(
    -child_neg_events_sum, #remove variable no longer needed after calculation
    -leq_b2_events_total_Y04,
    -leq_b2_neg_events_total_Y04
  ) %>% 
  mutate( #create cumulative adversity variable
     postnatal_stress = as.numeric(as.character(postnatal_stress)),
     sum_advrsty = maltreated_mod_rev + stai_state_above_clin_cutoff_pw26 + postnatal_stress,
     cumulative_adv = case_when(
      sum_advrsty == 0 ~ 0,
      sum_advrsty == 1 ~ 1,
      sum_advrsty == 2 | sum_advrsty == 3 ~ 2
    )
  )

#create M24 metadata file, standardize cbcl scores within sex 
 metadata_M24 <-
  metadata_final %>% 
  dplyr::filter(Timepoint_months==24) %>% 
  group_by(sex_binary) %>% 
  dplyr::mutate(across(contains("tot_"), .fns = list(std = ~as.numeric(scale(.))*15 + 100), .names="{col}_{fn}")) %>% 
  ungroup()

#remove unused cbcl variables  
metadata_M24 <- metadata_M24 %>% 
  dplyr::select(-(!contains("_std") & contains("tot_")))

#create separate dataframe for M48 metadata
metadata_Y4 <-
  metadata_final %>% 
  filter(Timepoint_months ==48)
```

# Write metadata file
## Write intial metadata file 
```{r}
metadata_output_path <- "../../data/microbiome/M24_M48/input_files/M24_metadata_typeheader1_rev_dqi.tsv"

write_tsv(metadata_M24, metadata_output_path)
```

## Manually add typeheader to metadata file
This is necessary for reading the metadata file into phyloseq (otherwise it will not recognize the file as metadata)  

1. Open metadata .tsv in Excel
2. Add row at top with #q2:types under the #SampleID column, and then for each column write either "categorical" or "numeric" depending on data type
3. Save as .txt

## Convert manually edited metadata file (typeheader added) to .tsv
```{r}
test_file <- "../../data/microbiome/M24_M48/input_files/M24_metadata_typeheader1_rev_dqi.txt"

test <- read_tsv(test_file)

write_tsv(test, "../../data/microbiome/M24_M48/input_files/M24_metadata_typeheader_new_rev_dqi.tsv")
```



